Towards a Verified Tableau Prover for a Ground Fragment of Set Theory

### 2023-02-15 paper/CADE/paper.tex
Continue

\section{Introduction}
In contrast to pen-and-paper proofs where we expect the reader to fill in trivial details, interactive theorem provers (ITPs) require us to justify each proof step rigorously.
\todo[inline]{Reviewer: Reformulate below because we can use cuts}
Manually reducing each proof step down to the axioms would be impractical, though;
hence, ITPs usually offer a wide array of automated methods to deal with the proof goals that arise during interactive use.

@@

We remark that $\longleftrightarrow$ is equivalent to \lstinline!$=$! on the type of Booleans \lstinline!bool! and \lstinline!$\equiv$! is definitional equality of the meta-logic of Isabelle/HOL, which is called Isabelle/Pure.
Meta-implication is denoted by \lstinline!$\Longrightarrow$! and a chain of implications
\begin{lstlisting}
  A$_\text{1}$ $\Longrightarrow$ $\cdots$ $\Longrightarrow$ A$_\text{k}$ $\Longrightarrow$ C
\end{lstlisting}
can be abbreviated by 
\begin{lstlisting}
  $\llbracket$ A$_\text{1}$;$\,\ldots\,$;A$_\text{k}$ $\rrbracket$ $\Longrightarrow$ C.
\end{lstlisting}

@@

Like the function \lstinline!vars!, the automatically defined function \lstinline!atoms :: 'a fm $\Rightarrow$ 'a set! retrieves all set atoms in a formula.
We combine these functions to extract all the variables occurring in a set formula.
\begin{lstlisting}
definition vars :: 'a pset_fm $\Rightarrow$ 'a set where
  vars $\phi$ $\equiv$ $\bigcup$(vars ` atoms $\phi$)
\end{lstlisting}

Similarly to \lstinline!vars!, we introduce the overloaded constant \lstinline!subterms :: 'b $\Rightarrow$ 'a pset_term set!, which returns all the set terms that are subterms of a set term, set atom, or formula, respectively. Lastly, we consider the subformulae of a formula and define a function \lstinline!subfms :: 'a fm $\Rightarrow$ 'a fm set! that computes them.
The definition of \lstinline!subterms! and \lstinline!subfms! is standard so we omit them here and refer to the appendix.

@@

\subsection{Semantics}
The original paper~\cite{new_fast_tableau} bases the semantics of \MLSS{} on the von Neumann hierarchy of sets $\mathcal{V}$. 
While there is an entry~\cite{zfc_in_hol_afp} in the \textit{Archive of Formal Proofs} (AFP) that axiomatises this hierarchy,
we instead use the hierarchy of \textit{hereditarily finite sets} (HF sets) which fulfil all the same axioms as $\mathcal{V}$ except for the axiom of infinity.
The HF sets are sufficient for our purposes as \MLSS{} is not powerful enough to force infinite models~\cite[Chapter~14.2]{set_theory}.
In contrast to $\mathcal{V}$, the HF sets are directly representable in Isabelle/HOL, and indeed, there is an AFP entry~\cite{hf_AFP} that formalises them.
The entry defines a type \lstinline!hf! that comes with the following functionality:
\begin{itemize}
  \item The function \lstinline!HF :: hf set $\Rightarrow$ set! that converts a finite set of HF sets into an HF set.
\item The usual set operations such as equality ($=$), membership ($\hfmem$), union ($\sqcup$), intersection ($\sqcap$), and difference ($-$) are defined.
\item Finally, the empty set coincides with the ordinal $0$, so it is denoted by \lstinline!0 :: hf!.
\end{itemize}

Equipped with the above, we define the interpretation functions
\begin{lstlisting}
  &\Ist& :: ('a $\Rightarrow$ hf) $\Rightarrow$ 'a pset_term $\Rightarrow$ hf &\textnormal{and}&
  &\Isa& :: ('a $\Rightarrow$ hf) $\Rightarrow$ 'a pset_atom $\Rightarrow$ hf
\end{lstlisting}
in the standard way, i.e.\ by mapping each syntactic construct to the corresponding operation on HF sets and interpreting variables with respect to a given valuation function \lstinline!M :: 'a $\Rightarrow$ hf!.
For the concrete definition we refer to the appendix.

@@

\section{A Tableau Calculus for MLSS}
We formalise the tableau calculus for \MLSS{} as described by \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}.
Inspired by the formalisation of a tableau calculus for hybrid logic by \citeauthor{hybrid_logic_afp}~\cite{hybrid_logic_afp}, we simply use a list to represent a branch of the tableau tree.
Note that formulae are added to the front of the list during branch expansion, so \lstinline!last b! for a branch \lstinline!b! is always the formula that we are trying to disprove with the tableau. 
\begin{lstlisting}
type_synonym 'a branch = 'a pset_fm list
\end{lstlisting}
We now lift the functions that extract the variables respectively subterms of formulae to branches.
\begin{lstlisting}
definition vars :: 'a branch $\Rightarrow$ 'a set 
  vars b $\equiv$ $\bigcup$(vars ` set b)

definition subterms :: 'a branch $\Rightarrow$ 'a pset_term set where
  subterms b $\equiv$ $\bigcup$(subterms ` set b)
\end{lstlisting}
In the standard tableau calculus for propositional logic as \citeauthor{tableau}~\cite{tableau} describes it, a branch is called \textit{closed} if it contains both the negation of a formula and the formula itself;
conversely, it is called \textit{open} if it is not closed.
For \MLSS{}, we extend the notion of closedness with three additional rules; the first two are straightforward while the last one states that a branch is closed when the branch contains a membership cycle
\begin{lstlisting}
  t$_\text{0}$ $\inS$ t$_\text{1}$, t$_\text{1}$ $\inS$ t$_\text{2}$, $\ldots$, t$_\text{k}$ $\inS$ t$_\text{0}$.
\end{lstlisting}

@@

\subsection{Linear Expansion Rules}
\begin{table*}[t]
  \centering
  \lstset{
    escapeinside={@}{@}
  }
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \begin{tabular}{>{\centering\arraybackslash}p{0.48\textwidth}m{0.02\textwidth}>{\centering\arraybackslash}p{0.48\textwidth}}
    \toprule
    Propositional Rules & & Rules for \lstinline!$\unionS$! \\
    \cmidrule{1-1}\cmidrule{3-3}
    \begin{tabular}{p{0.22\textwidth}m{0.06\textwidth}p{0.2\textwidth}}
      \lstinline!@\fmAnd{p}{q}@! & $\Longrightarrow$ & \lstinline!p, q! \\
      \lstinline!@\fmNeg{(\fmOr{p}{q})}@! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@, @\fmNeg{q}@! \\
      \lstinline!@\fmOr{p}{q}, \fmNeg{p}@! & $\Longrightarrow$ & \lstinline!q! \\
      \lstinline!@\fmOr{p}{q}, \fmNeg{q}@! & $\Longrightarrow$ & \lstinline!p! \\
      \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, p! & $\Longrightarrow$ & \lstinline!@\fmNeg{q}@! \\
      \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, q! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@! \\
      \lstinline!@\fmNeg{(\fmNeg{p})}@! & $\Longrightarrow$ & \lstinline!p!\\[6ex]
    \end{tabular}
                        & &
    \begin{tabular}{p{0.22\textwidth}m{0.06\textwidth}p{0.2\textwidth}}
    \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$!, \newline \lstinline!s $\notinS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!s $\in$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ton$! & ~\newline $\Longrightarrow$ & ~\newline \lstinline!s $\inS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$! \\
    \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! \\
    \end{tabular}
    \\
    \\

    Rules for \lstinline!Single! && Rules for \lstinline!$\eqS$! \\
    \cmidrule{1-1} \cmidrule{3-3}
    \begin{tabular}{p{0.22\textwidth}m{0.06\textwidth}p{0.2\textwidth}}
      & $\Longrightarrow$ & \lstinline!s $\inS$ Single s! \\
      \lstinline!s $\inS$ Single t! & $\Longrightarrow$ & \lstinline!s $\eqS$ t! \\
      \lstinline!s $\notinS$ Single t! & $\Longrightarrow$ & \lstinline!s $\neqS$ t! \\
    \end{tabular}
    & &
    \begin{tabular}{p{0.22\textwidth}m{0.06\textwidth}p{0.2\textwidth}}
      \lstinline!$\ton$ $\eqS$ $\ttw$, l! & $\Longrightarrow$ & \lstinline!l{$\ttw$/$\ton$}! \\
      \lstinline!$\ton$ $\eqS$ $\ttw$, l! & $\Longrightarrow$ & \lstinline!l{$\ton$/$\ttw$}! \\
      \lstinline!s$_\text{1}$ $\inS$ t, s$_\text{2}$ $\notinS$ t! & $\Longrightarrow$ & \lstinline!s$_\text{1}$ $\neqS$ s$_\text{2}$! \\
    \end{tabular}
    \\
    \bottomrule
  \end{tabular}
  \bigskip
  \caption{Linear expansion rules.\label{tab:lexpands}}
\end{table*}
The calculus considers two kinds of branch expansion rules: \textit{linear} and \textit{branching} rules.
As the name suggests, branching rules lead to the creation of new branches in the tableau while linear rules only extend a branch \lstinline!b! with new formulae \lstinline[breaklines=true]!b' = [$\psi_1$,$\ldots$,$\psi_n$]!, which we denote by \lstinline!&\lexpands{b'}{b}&!.
\autoref{tab:lexpands} shows the linear expansion rules.
Note that in the first two rules for \lstinline!$\eqS$!, \lstinline!l! is a literal occurring in the branch.
Furthermore, the term-for-term substitution \lstinline!l{s/t}! is restricted to the top-level set terms of \lstinline!l!, i.e.\ the set terms that occur directly under one of the atom constructors \lstinline!$\inS$! or \lstinline!$\eqS$!;
for example, given the literal
\begin{lstlisting}
  l = &\fmNeg{((s $\unionS$ u) $\diffS$ s $\eqS$ s $\unionS$ u)}&
\end{lstlisting}
we have
\begin{lstlisting}
  &\phantom{=}& (&\fmNeg{((s $\unionS$ u) $\diffS$ s $\eqS$ s $\unionS$ u)}&){t/s $\unionS$ u}
  = &\fmNeg{((s $\unionS$ u) $\diffS$ s $\eqS$ t)}&.
\end{lstlisting}
The perhaps more crucial restriction of the linear rules is that that no new subterm may be created by their application;
for instance, the second rule for \lstinline!$\unionS$!
\begin{lstlisting}
  s $\inS$ t$_\text{1}$ $\Longrightarrow$ s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$,
\end{lstlisting}
which formally represents 
\begin{lstlisting}
  (s $\inS$ t$_\text{1}$) $\in$ set b $\Longrightarrow$ &\lexpands{[s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$]}{b}&,
\end{lstlisting}
may only be used under the condition
\begin{lstlisting}
  t$_\text{1}$ $\unionS$ t$_\text{2}$ $\in$ subterms (last b).
\end{lstlisting}
The purpose of this restriction is to prevent unbounded expansion of the branch.
In fact, we give an explicit upper bound for the number of formulae in a branch in \autoref{sec:bound}.

@@

Finally, we remark that the original paper~\cite{new_fast_tableau} is missing the last propositional rule dealing with double negation.
This rule is required for completeness, though, considering that the branch
\begin{lstlisting}
  [&\fmNeg{\fmNeg{\fmNeg{p}}}&, p, &\fmAnd{\fmNeg{\fmNeg{\fmNeg{p}}}}{p}&]
\end{lstlisting}
is saturated --- neither linear nor branching rules apply --- and open, but there clearly is no model for the initial formula \lstinline!&\fmAnd{\fmNeg{\fmNeg{\fmNeg{p}}}}{p}&!.

@@

@@

Note that even branching rules are defined such that they never create new subterms with exception of the last rule, which adds a new variable to the branch.
These variables serve to manifest an inequality; hence, we call them \textit{witnesses}.
\begin{lstlisting}
definition wits :: 'a branch $\Rightarrow$ 'a set where
  wits b $\equiv$ vars b - vars (last b)
\end{lstlisting}

\section{A Decision Procedure for MLSS}
\begin{lstlisting}
function (domintros) mlss_proc_branch :: 'a branch $\Rightarrow$ bool where
  $\neg$ lin_sat b $\Longrightarrow$ mlss_proc_branch b =
  mlss_proc_branch ((SOME b'. &\lexpands{b'}{b}& $\land$
                              set b $\subset$ set (b' @ b)) @ b)
| $\llbracket$ lin_sat b; bclosed b $\rrbracket$ $\Longrightarrow$ mlss_proc_branch b = True
| $\llbracket$ $\neg$ sat b; bopen b; lin_sat b $\rrbracket$ $\Longrightarrow$ mlss_proc_branch b =
  ($\forall$b' $\in$ (SOME bs. &\fexpands{bs}{b}&). mlss_proc_branch (b' @ b))
| $\llbracket$ lin_sat b; sat b $\rrbracket$ $\Longrightarrow$ mlss_proc_branch b = bclosed b

@@

We abstractly specify the behaviour of the decision procedure, leaving an executable specification from which code can be generated to future work.
The implementation as displayed in TODO uses a couple of features of Isabelle/HOL's function package:
instead of defining the function via pattern matching, we specify the equations of the function as conditional rewrite rules.
This requires us to prove that the assumptions of the equations are non-overlapping, which is done by automation.
The other concern is that Isabelle/HOL requires functions to be total, so a recursive function needs to terminate in order for it to be well-defined;
nevertheless, the termination proof is separated from the definition of the function for modularity.
The function package maintains the soundness of the definition by introducing a so-called domain predicate \lstinline!mlss_proc_branch_dom! which characterises for which arguments the function terminates.
Each equation of the function is guarded by an assumption that the predicate holds for the argument.
To be able to reason about this predicate, we pass the \lstinline!domintros! flag that instructs the function package to derive introduction rules for it.
In \autoref{sec:correct}, we will use these rules to discharge the guarding assumptions for the context that \lstinline!mlss_proc_branch! is called in.
Before we go into more detail on how the termination is proved, we discuss its definition.

@@

Again, we need to ensure that the assumptions of the equations are non-overlapping and that the function terminates.
The former is taken care of by automation, leaving us to prove termination.
The assumption that \lstinline!b! is open implies that there are no membership cycles and thus \lstinline!bgraph b! is acyclic.
TODO: Shorten this proof.
Now, consider the ancestors \lstinline!av! of a vertex \lstinline!v!, which are those vertices from which we can reach \lstinline!v!.
Due to acyclicity, it holds that the ancestors \lstinline!ap! of a parent \lstinline!p! of \lstinline!v! are a subset of \lstinline!av!.
They are also a proper subset of \lstinline!av! since \lstinline!p! is an ancestor of \lstinline!v! but not of \lstinline!p! itself.
Together with the fact that \lstinline!bgraph b! is finite, it follows that the cardinality of the ancestors decreases in each recursive call, thus proving the termination of \lstinline!realise!.

@@

\subsection{Characterisation of the Pure Witnesses\label{sec:pwits}}
Recall that the pure witnesses of a branch \lstinline!b! are those witnesses that are not related to other subterms in \lstinline!last b! by equality.
In the context of a well-formed branch, this characterisation can be strengthened to any set term and, in addition, we also get that there is no membership literal where a pure witness is on the right-hand side.
Intuitively speaking, the realisation of a pure witness does not depend on the realisation of any other set term.
\begin{lstlisting}[label={lst:lemma_2}]
lemma lemma_2:
  assumes wf_branch b and c $\in$ pwits b
  shows (Var c $\eqS$ t) $\notin$ set b and (t $\eqS$ Var c) $\notin$ set b
    and (t $\inS$ Var c) $\notin$ set b
\end{lstlisting}
So why are pure witnesses treated differently?
According to the definition of \lstinline!realise!, the pure witnesses would be evaluated to the empty set \lstinline!0 :: hf!, were they not treated separately. 
To see that this is a problem, consider the branch
\begin{lstlisting}
  b = [Var s $\neqS$ Var t, Var t $\neqS$ Var u]
\end{lstlisting}
which expands to several open and saturated branches, one of which is
\begin{lstlisting}
  [Var x $\neqS$ Var y,
   Var x $\inS$ Var s, Var x $\notinS$ Var t,
   Var y $\inS$ Var t, Var y $\notinS$ Var u] @ b
\end{lstlisting}
for some fresh \lstinline!x! and \lstinline!y!.
Assigning both \lstinline!Var x! and \lstinline!Var y! a value of \lstinline!0! would contradict the literal \lstinline!Var x $\neqS$ Var y!.
To prevent this, we assign the pure witnesses pairwise different values.

@@

\noindent The above lemma allows us to use the computation induction rule of \lstinline!mlss_proc_branch! on well-formed branches, which we do to prove soundness and completeness, starting with completeness.
\begin{lstlisting}[belowskip=0pt]
lemma mlss_proc_branch_complete:
  fixes b :: 'a branch
  assumes wf_branch b and $\neg$ mlss_proc_branch b
  assumes infinite (UNIV :: 'a set)
  shows $\exists$M. M $\models$ last b
\end{lstlisting}
\begin{proof}
  We proceed to prove the completeness by computation induction on \lstinline!b!.

  The cases of the first and third equation, where the branch can be expanded, are immediately resolved with the induction hypothesis.

  We evaluate \lstinline!mlss_proc_branch b! to \lstinline!True! in the case of the third equation, yielding a contradiction with the assumption \lstinline!$\neg$ mlss_proc_branch b!.

>>>>

\section{Introduction}
In contrast to pen-and-paper proofs where we expect the reader to fill in trivial details, interactive theorem provers (ITPs) require us to justify each proof step rigorously.
Manually reducing each proof step down to the axioms would be impractical, though;
hence, ITPs usually offer a wide array of automated methods to deal with the proof goals that arise during interactive use.

@@

We remark that $\longleftrightarrow$ is equivalent to \lstinline!$=$! on the type of Booleans \lstinline!bool! and \lstinline!$\equiv$! is definitional equality of the meta-logic of Isabelle/HOL, which is called Isabelle/Pure.
Meta-implication is denoted by \lstinline!$\Longrightarrow$! and a chain of implications
\lstinline!A$_\text{1}$ $\Longrightarrow$ $\cdots$ $\Longrightarrow$ A$_\text{k}$ $\Longrightarrow$ C!
can be abbreviated by 
\lstinline!$\llbracket$ A$_\text{1}$;$\,\ldots\,$;A$_\text{k}$ $\rrbracket$ $\Longrightarrow$ C!.

@@

Like the function \lstinline!vars!, the automatically defined function \lstinline!atoms :: 'a fm $\Rightarrow$ 'a set! retrieves all set atoms in a formula.
We combine these functions to extract all the variables occurring in a set formula.
\begin{lstlisting}
definition vars $\phi$ $\equiv$ $\bigcup$(vars ` atoms $\phi$)
\end{lstlisting}

Similarly to \lstinline!vars!, we introduce the overloaded constant \lstinline!subterms :: 'b $\Rightarrow$ 'a pset_term set!, which returns all the set terms that are subterms of a set term, set atom, or formula, respectively. Lastly, we consider the subformulae of a formula and define a function \lstinline!subfms :: 'a fm $\Rightarrow$ 'a fm set! that computes them.
The functions \lstinline!subterms! and \lstinline!subfms! are defined in the usual way (cf. Appendix TODO).

@@

\subsection{Semantics}
The original paper~\cite{new_fast_tableau} bases the semantics of \MLSS{} on the von Neumann hierarchy of sets $\mathcal{V}$. 
While there is an entry~\cite{zfc_in_hol_afp} in the \textit{Archive of Formal Proofs} (AFP) that axiomatises this hierarchy,
we instead use the hierarchy of \textit{hereditarily finite sets} (HF sets) which fulfil all the same axioms as $\mathcal{V}$ --- that is, the axioms of ZF --- except for the axiom of infinity.\todo{Membership is well-founded.}
The HF sets are sufficient for our purposes as \MLSS{} is not powerful enough to force infinite models~\cite[Chapter~14.2]{set_theory}.
In contrast to $\mathcal{V}$, the HF sets are directly representable in Isabelle/HOL, and indeed, there is an AFP entry~\cite{hf_AFP} that formalises them.
The entry defines a type \lstinline!hf! that comes with the following functionality:
\begin{itemize}
  \item The function \lstinline!HF :: hf set $\Rightarrow$ set! that converts a finite set of HF sets into an HF set.
\item The usual set operations such as equality ($=$), membership ($\hfmem$), union ($\sqcup$), intersection ($\sqcap$), and difference ($-$) are defined.
\item Finally, the empty set coincides with the ordinal $0$, so it is denoted by \lstinline!0 :: hf!.
\end{itemize}

Equipped with the above, we define the interpretation functions
\begin{itemize}
  \item \lstinline!&\Ist& :: ('a $\Rightarrow$ hf) $\Rightarrow$ 'a pset_term $\Rightarrow$ hf! and 
  \item \lstinline!&\Isa& :: ('a $\Rightarrow$ hf) $\Rightarrow$ 'a pset_atom $\Rightarrow$ hf!
\end{itemize}
in the standard way, i.e.\ by mapping each syntactic construct to the corresponding operation on HF sets and interpreting variables with respect to a given valuation function \lstinline!M :: 'a $\Rightarrow$ hf!.
For the concrete definition we refer to the appendix.

@@

\section{A Tableau Calculus for MLSS}
We formalise the tableau calculus for \MLSS{} as described by \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}.
Inspired by the formalisation of a tableau calculus for hybrid logic by \citeauthor{hybrid_logic_afp}~\cite{hybrid_logic_afp}, we simply use a list to represent a branch of the tableau tree.
Note that formulae are added to the front of the list during branch expansion, so \lstinline!last b! for a branch \lstinline!b! is always the formula that we are trying to disprove with the tableau. 
\begin{lstlisting}
type_synonym 'a branch = 'a pset_fm list
\end{lstlisting}

\noindent The functions \lstinline!vars! and \lstinline!subterms! are lifted to branches in the expected way.

In the standard tableau calculus for propositional logic as \citeauthor{tableau}~\cite{tableau} describes it, a branch is called \textit{closed} if it contains both the negation of a formula and the formula itself;
conversely, it is called \textit{open} if it is not closed.
For \MLSS{}, we extend the notion of closedness with three additional rules; the first two are straightforward while the last one states that a branch is closed when the branch contains a membership cycle
\lstinline!t$_\text{0}$ $\inS$ t$_\text{1}$, t$_\text{1}$ $\inS$ t$_\text{2}$, $\ldots$, t$_\text{k}$ $\inS$ t$_\text{0}$!.

@@

\subsection{Linear Expansion Rules}
\begin{table*}[t]
  \centering
  \lstset{
    escapeinside={@}{@}
  }
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \begin{tabular}{ccc}
    \toprule
    Propositional Rules & {\hskip 6pt} & Rules for \lstinline!$\unionS$! \\
    \cmidrule(l{1pt}){1-1}\cmidrule(r{1pt}){3-3}
    \begin{tabular}{p{0.18\textwidth}m{0.05\textwidth}p{0.175\textwidth}}
      \lstinline!@\fmAnd{p}{q}@! & $\Longrightarrow$ & \lstinline!p, q! \\
      \lstinline!@\fmNeg{(\fmOr{p}{q})}@! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@, @\fmNeg{q}@! \\
      \lstinline!@\fmOr{p}{q}, \fmNeg{p}@! & $\Longrightarrow$ & \lstinline!q! \\
      \lstinline!@\fmOr{p}{q}, \fmNeg{q}@! & $\Longrightarrow$ & \lstinline!p! \\
      \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, p! & $\Longrightarrow$ & \lstinline!@\fmNeg{q}@! \\
      \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, q! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@! \\
      \lstinline!@\fmNeg{(\fmNeg{p})}@! & $\Longrightarrow$ & \lstinline!p!\\[6ex]
    \end{tabular}
                        & &
    \begin{tabular}{p{0.22\textwidth}m{0.05\textwidth}p{0.22\textwidth}}
    \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!s $\in$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$! \\
    \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! \\
    \end{tabular}
    \\
    \\

    Rules for \lstinline!$\interS$! && Rules for \lstinline!$\diffS$! \\
    \cmidrule(l{1pt}){1-1}\cmidrule(r{1pt}){3-3}
    \begin{tabular}{c}
      $\vdots$ \\
    \end{tabular}
    & &
    \begin{tabular}{c}
      $\vdots$ \\
    \end{tabular}
    \\
    \\

    Rules for \lstinline!Single! && Rules for \lstinline!$\eqS$! \\
    \cmidrule(l{1pt}){1-1}\cmidrule(r{1pt}){3-3}
    \begin{tabular}{p{0.18\textwidth}m{0.05\textwidth}p{0.175\textwidth}}
      & $\Longrightarrow$ & \lstinline!s $\inS$ Single s! \\
      \lstinline!s $\inS$ Single t! & $\Longrightarrow$ & \lstinline!s $\eqS$ t! \\
      \lstinline!s $\notinS$ Single t! & $\Longrightarrow$ & \lstinline!s $\neqS$ t! \\
    \end{tabular}
    & &
    \begin{tabular}{p{0.22\textwidth}m{0.05\textwidth}p{0.22\textwidth}}
      \lstinline!$\ton$ $\eqS$ $\ttw$, l! & $\Longrightarrow$ & \lstinline!l{$\ttw$/$\ton$}! \\
      \lstinline!$\ton$ $\eqS$ $\ttw$, l! & $\Longrightarrow$ & \lstinline!l{$\ton$/$\ttw$}! \\
      \lstinline!s$_\text{1}$ $\inS$ t, s$_\text{2}$ $\notinS$ t! & $\Longrightarrow$ & \lstinline!s$_\text{1}$ $\neqS$ s$_\text{2}$! \\
    \end{tabular}
    \\
    \bottomrule
  \end{tabular}
  \bigskip
  \caption{Linear expansion rules.\label{tab:lexpands}}
\end{table*}
The calculus considers two kinds of branch expansion rules: \textit{linear} and \textit{branching} rules.
As the name suggests, branching rules lead to the creation of new branches in the tableau while linear rules only extend a branch \lstinline!b! with new formulae \lstinline[breaklines=true]!b' = [$\psi_1$,$\ldots$,$\psi_n$]!, which we denote by \lstinline!&\lexpands{b'}{b}&!.
\autoref{tab:lexpands} shows the linear expansion rules.
Note that in the first two rules for \lstinline!$\eqS$!, \lstinline!l! is a literal occurring in the branch.
Furthermore, the term-for-term substitution \lstinline!l{s/t}! is restricted to replace \lstinline!t! only if it occurs as a top-level set terms of \lstinline!l!, i.e.\ it occurs directly under one of the atom constructors \lstinline!$\inS$! or \lstinline!$\eqS$!.

The more crucial restriction of the linear rules is that that no new subterm may be created by their application;
for instance, the second rule for \lstinline!$\unionS$! is
\lstinline!s $\inS$ t$_\text{1}$ $\Longrightarrow$ s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$!,
which formally represents 
\begin{lstlisting}
  (s $\inS$ t$_\text{1}$) $\in$ set b $\Longrightarrow$ &\lexpands{[s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$]}{b}&,
\end{lstlisting}
may only be used under the condition
\lstinline[breaklines]!t$_\text{1}$ $\unionS$ t$_\text{2}$ $\in$ subterms (last b)!.
The purpose of this restriction is to prevent unbounded expansion of the branch.
In fact, we give an explicit upper bound for the number of formulae in a branch in \autoref{sec:bound}.

@@

Finally, we remark that the original paper~\cite{new_fast_tableau} is missing the last propositional rule dealing with double negation.
This rule is required for completeness, though, considering that the branch
\lstinline![&\fmNeg{\fmNeg{\fmNeg{p}}}&, p, &\fmAnd{\fmNeg{\fmNeg{\fmNeg{p}}}}{p}&]!
is saturated --- neither linear nor branching rules apply --- and open, but there clearly is no model for the initial formula \lstinline!&\fmAnd{\fmNeg{\fmNeg{\fmNeg{p}}}}{p}&!.

@@

      $\vdots$ & $\vdots$ & $\vdots$ \\

      \addlinespace[0.6em]

@@

Note that even branching rules are defined such that they never create new subterms with exception of the last rule, which adds a new variable to the branch.
These variables serve to manifest an inequality; hence, we call them \textit{witnesses}.
\begin{lstlisting}
definition wits b $\equiv$ vars b - vars (last b)
\end{lstlisting}

\section{A Decision Procedure for MLSS}
\begin{lstlisting}
function mlss_proc_branch :: 'a branch $\Rightarrow$ bool where
  $\neg$ lin_sat b $\Longrightarrow$ mlss_proc_branch b =
  mlss_proc_branch ((SOME b'. &\lexpands{b'}{b}& $\land$
                              set b $\subset$ set (b' @ b)) @ b)
| $\llbracket$ lin_sat b; bclosed b $\rrbracket$ $\Longrightarrow$ mlss_proc_branch b = True
| $\llbracket$ $\neg$ sat b; bopen b; lin_sat b $\rrbracket$ $\Longrightarrow$ mlss_proc_branch b =
  ($\forall$b' $\in$ (SOME bs. &\fexpands{bs}{b}&). mlss_proc_branch (b' @ b))
| $\llbracket$ lin_sat b; sat b $\rrbracket$ $\Longrightarrow$ mlss_proc_branch b = bclosed b

@@

We abstractly specify the behaviour of the decision procedure, leaving an executable specification from which code can be generated to future work.
The implementation as displayed in TODO uses a couple of features of Isabelle/HOL's function package:
instead of defining the function via pattern matching, we specify the equations of the function as conditional rewrite rules.
This requires us to prove that the assumptions of the equations are non-overlapping, which is done by automation.
The other concern is that Isabelle/HOL requires functions to be total, so a recursive function needs to terminate in order for it to be well-defined;
nevertheless, the termination proof is separated from the definition of the function for modularity.
The function package maintains the soundness of the definition by introducing a so-called domain predicate \lstinline!mlss_proc_branch_dom! which characterises for which arguments the function terminates.
Each equation of the function is guarded by an assumption that the predicate holds for the argument.
In \autoref{sec:correct}, we will show that the domain predicate holds for the context that \lstinline!mlss_proc_branch! is called in.
Before we go into more detail on how the termination is proved, we discuss its definition.

@@

Again, we need to ensure that the assumptions of the equations are non-overlapping and that the function terminates.
The former is taken care of by automation, leaving us to prove termination.
The assumption that \lstinline!b! is open implies that there are no membership cycles and thus \lstinline!bgraph b! is acyclic.
Furthermore, the graph is finite by definition.
Thus, we can use the cardinality of the set of ancestors as a measure that decreases in each recursive call.

@@

\subsection{Characterisation of the Pure Witnesses\label{sec:pwits}}
Recall that the pure witnesses of a branch \lstinline!b! are those witnesses that are not related to other subterms in \lstinline!last b! by equality.
In the context of a well-formed branch, this characterisation can be strengthened to any set term and, in addition, we also get that there is no membership literal where a pure witness is on the right-hand side.
Intuitively speaking, the realisation of a pure witness does not depend on the realisation of any other set term.
\begin{lstlisting}[label={lst:lemma_2}]
lemma lemma_2:
  assumes wf_branch b and c $\in$ pwits b
  shows (Var c $\eqS$ t) $\notin$ set b and (t $\eqS$ Var c) $\notin$ set b
    and (t $\inS$ Var c) $\notin$ set b
\end{lstlisting}
So why are pure witnesses treated differently?
According to the definition of \lstinline!realise!, the pure witnesses would be evaluated to the empty set \lstinline!0 :: hf!, were they not treated separately. 
To see that this is a problem, consider the branch
\lstinline!b = [Var s $\neqS$ Var t, Var t $\neqS$ Var u]!
which expands to several open and saturated branches, one of which is
\begin{lstlisting}
  [Var x $\neqS$ Var y, Var x $\inS$ Var s, Var x $\notinS$ Var t,
  &\phantom{[Var x $\neqS$ Var y,}& Var y $\inS$ Var t, Var y $\notinS$ Var u] @ b
\end{lstlisting}
for some fresh \lstinline!x! and \lstinline!y!.
Assigning both \lstinline!Var x! and \lstinline!Var y! a value of \lstinline!0! would contradict the literal \lstinline!Var x $\neqS$ Var y!.
To prevent this, we assign the pure witnesses pairwise different values.

@@

\noindent The above lemma allows us to use the computation induction rule of \lstinline!mlss_proc_branch! on well-formed branches, which we use to prove soundness and completeness.
As both proofs are essentially an application of soundness, respectively completeness, of the calculus, we refer the reader to the formalisation.
\begin{lstlisting}
lemma mlss_proc_branch_complete:
  fixes b :: 'a branch
  assumes wf_branch b and $\neg$ mlss_proc_branch b
  assumes infinite (UNIV :: 'a set)
  shows $\exists$M. M $\models$ last b

lemma mlss_proc_branch_sound:
  assumes wf_branch b and $\forall$$\psi$ $\in$ set b. M $\models$ $\psi$
  shows $\neg$ mlss_proc_branch b
\end{lstlisting}
